A subjective Bayesian framework for synthesizing deep uncertainties in climate risk management

### 2022-07-02 papers/earths-future/submission.tex
incorporate james and adam proofreading

\begin{keypoints}
  \item Decisions about how to manage climate risks are often made under deep uncertainty.
  \item We introduce a Bayesian framework to transparently synthesize and characterize deep uncertainties with the goal to support decision-making
  \item We demonstrate the framework using a simple case study of house elevation for coastal flood risk management
\end{keypoints}

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT and PLAIN LANGUAGE SUMMARY
%
% A good Abstract will begin with a short description of the problem
% being addressed, briefly describe the new data or analyses, then
% briefly states the main conclusion(s) and how they are supported and
% uncertainties.

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have 
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%

%% \begin{abstract} starts the second page

\begin{abstract}
  Projections of future climate risks can vary considerably between sources without consensus regarding their relative probability.
  This deep uncertainty poses considerable communication and decision-analytical challenges.
  One such challenge is how to present trade-offs under deep uncertainty in a salient and interpretable manner.
  Some common approaches include analyzing a small subset of projections or invoking Laplace's principle of insufficient reason to justify a simple average.
  These approaches can underestimate risks, hide deep uncertainties, and provide little insights into which assumptions drive decision-relevant outcomes.
  Here we introduce and demonstrate a transparent Bayesian framework for synthesizing deep uncertainties to inform decision support.
  The first step of this workflow is to create an ensemble of simulations representing possible futures and analyze them through standard exploratory modeling techniques.
  Next, a small set of probability distributions representing subjective beliefs over possible futures is used to weight the scenarios.
  Finally, these weights are used to compute and characterize trade-offs, conduct robustness checks, and reveal implicit assumptions.
  We demonstrate the framework through a didactic case study analyzing how high to elevate a house to manage coastal flood risks.
\end{abstract}

@@

What are sound strategies to manage risks driven by climatic changes?
Addressing this question is complicated by the large uncertainties surrounding projections of the coupled natural-human systems.
These uncertainties often arise from choices experts have to make, for example about technical detail on how to model future water levels.
Different experts can disagree about these choices, leading to different plausible projections.
Analyzing decisions in such a situation of deep uncertainty poses nontrivial challenges.
At one extreme, just picking a single representative projection can lead to under-estimation of risk and poor decisions.
At the other extreme, communicating results separately for each projection can overwhelm decision-makers.
To make matters worse, typical approaches to this problem are mostly silent on what assumptions make a difference for the decisions at hand.
We develop and demonstrate a framework to address these challenges.
The framework provides a transparent approach to (i) combine a large number of deeply uncertain projections to a more understandable and smaller sample and (ii) provide insights as to how assumptions and modeling choices made by experts influence decisions.
We demonstrate the approach with a relatively simple example question of how high to elevate a house in the face of deeply uncertain projections about future water levels.

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

@@

Many critical infrastructure services around the world are aging and inadequate, and anticipated changes to regulation, economics and finance, patterns of population and infrastructure use, and climate challenge their ability to meet societal needs \cite<\eg,>[]{doss-gollin_txtreme:2021,doss-gollin_fatalism:2020,chester_reliable:2020,ho_dams:2017}.
The need to upgrade existing infrastructure and build new facilities motivates the question: for which possible futures should infrastructure systems and components be designed?
The answer to this question depends in part on values: intrinsic trade-offs between safety, performance, cost, and other objectives depend on context and stakeholder preferences \cite{keller_management:2021} and are often regulated by statue or industry guidelines.
For example, hospitals and critical infrastructure are generally designed to a higher standard of risk protection than ordinary buildings.
At the same time, performance depends upon future conditions, and so decisions about which scenario to design for are necessarily subject to implicit or explicit assumptions about the likelihood or possibility of different future conditions.

@@

Current practice in engineering, infrastructure design, and local governance relies heavily upon standards that specify particular design events or conditions that buildings and infrastructure should safely withstand \cite{asce_7-10:2013,bruneau_multihazard:2017}.
These are often, though not always, informed by probabilistic analysis of relevant data.
For example, \gls{fema}, local governments, and engineering consultants produce local floodplain maps in many communities.
These trigger specific floodplain regulations, such as a requirement for homes in the 100-year floodplain with federally backed mortgages to be covered by flood insurance \cite{kousky_voucher:2014}.
Additionally, local building codes \cite<based on guidance such as>[]{asce_24-05:2006,FEMA_p-55:2011} may require new construction in flood zones to be elevated freeboard above a nominal \gls{bfe}.
Design standards are also used to design large-scale infrastructure.
For example, some levees in the Netherlands are required to be designed such that the annual probability of failure is less than 1 in \num{4000} \cite{eijgenraam_flooding:2014}, while a seawall proposed as part of a \$29 billion coastal protection project for  Galveston Bay was designed by setting the nominal annual probability of overtopping to 1\% \cite[Appendix D.,~p.~2-59]{USACE_coastal:2021}.

@@

Currently methods for estimating the \glspl{pdf} of climate hazards used in standard-setting and decision-making emphasize nominally objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies procedures for estimating flood frequency \cite{bulletin17c:2019} and \gls{noaa} Atlas 14 provides estimates of the intensity, duration, and frequency of extreme rainfall \cite{atlas14_texas:2018}.
Among several statistical assumptions these analyses make that have been recently called into question is stationarity \cite<the assumption that past and future hazard come from the same \gls{pdf}; see>[for a review]{merz_review:2014}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \cite{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}, and trends in many other hazards are consistent with observations \cite<see>[for a comprehensive summary]{ipcc_impacts:2022}.
While some methods have been proposed for incorporating trends into these analyses \cite<see>[for a review]{Salas:2018ge}, these assume specific forms of nonstationarity which may not fully represent physical processes or true levels of uncertainty \cite{DossGollin:2019,Montanari:2014hl,Serinaldi:2015bq}.
Ultimately, the difficulty of incorporating nonstationarity into existing frameworks has led to continued reliance on stationarity despite recognition of the associated limitations as discussed in p.~2 of \citeA{bulletin17c:2019} and p.~A.4-42 of \citeA{atlas14_texas:2018}.

@@

Although these methods for \gls{dmdu} have proven valuable in a wide range of settings, they still require necessarily subjective assessments about the likelihood of future conditions.
Even exploratory models require the analyst to choose which uncertainties are considered and how they are sampled.
When analyses use metrics that integrate performance over many possible futures, whether to compute sensitivities, estimate expectations, or compute robustness metrics, they necessarily make assumptions about the likelihood of different futures.
For example, uniform priors are often justified by deferring to Laplace's principle of insufficient reason \cite<see>[p.~135]{stigler_uncertainty:1986}, but even this choice makes strong assumptions that are sensitive to factors such as parameterization \cite[p.54]{gelman_bda3:2014}.
While one interpretation of deep uncertainty is that that specifying a joint \gls{pdf} over inputs is inappropriate, assumptions about the ranges and independence of parameters to sample are just as subjective as the choice of probability distribution \cite{schneider_scenarios:2002,quinn_exploratory:2020} and, indeed, can be interpreted  as a specific choice of probability distribution (we revisit this point in \cref{sec:analysis-condition}).
This motivates the development of decision analytic frameworks that draw from the strengths of \gls{dmdu} methods such as exploratory modeling, vulnerability assessment, robustness checks, and iterative stakeholder critique, but that embrace the reality that assumptions about the future are inescapable.

@@

We are motivated by parallels between decision making under deep uncertainty and the statistical problem of model selection.
\citeA{oreskes_verification:1994} argues that because natural systems are never closed and model results are never unique, validation and verification of models representing these systems is necessarily qualitative and subjective.
In the literature on decision making under deep uncertainty, this has led to recognition of the need to develop strategies that are robust to model imperfections.
We find parallels to this challenge in the Bayesian literature on model selection, particularly in the ``$\mathcal{M}$-open'' case in which there is no ``true'' model to identify and all models are imperfect representations of reality.
This is a theoretical contrast to the ``$\mathcal{M}$''-closed case in which a true model can be identified and written down but is one amongst finitely many models from which an analyst has to choose as well as the ``$\mathcal{M}$''-complete case in which a true model exists but is inaccessible in the sense that even though it can be conceptualized it cannot be written down \cite{bernardo_bayesian:1994}.
Whereas model selection in the $\mathcal{M}$-closed case seeks to asymptotically identify the true model from a set of candidate models, model selection in the $\mathcal{M}$-open case emphasizes a subjective view, both of probability itself and of the modeling process, for which probability offers a self-consistent language with which to reason about the unknown rather than a statement of objective truth (see \citeA{gelman_philosophy:2013}, \citeA{ramsey_probability:2016}, or \citeA{jaynes_probability:2003} for a discussion of Bayesian philosophy and \citeA{Piironen:2017eh} for a technical discussion of methods for model selection).
From a practical perspective, model selection in the $\mathcal{M}$-open case emphasizes building models iteratively, simulating the consequences of these models, and subsequently using these simulations to critique and improve them \cite{gelman_workflow:2020}.

@@

We proceed as follows.
In \cref{sec:analysis} we present three formal decision analytic frameworks for analyzing an ensemble of \gls{slr} simulations, building through existing approaches for exploratory modeling (\cref{sec:analysis-explore}) and scenario analysis (\cref{sec:analysis-condition}) towards a formal Bayesian method for transparently synthesizing deep uncertainty through subjective prior beliefs.
In \cref{sec:case-study} we describe the case study in detail.
Next in \cref{sec:results} we present results for each of the three decision lenses and discuss the advantages and limitations of each theoretical approach.
Finally in \cref{sec:conclusions} we discuss key findings, future research needs, and implications for research, policy, and practice.

@@

A first analytical step is to use the model in an ``exploratory'' mode.
Exploratory modeling is averse to making explicit assumptions about the likelihood of different \glspl{sow} and instead seek to generate new knowledge \cite{bankes:1993} by systematically exploring a large number of possible futures, emphasizing interactions between different uncertainties \cite{reed_msdbook:2022}.
Exploratory modeling is often paired with analyses that identify relevant scenarios  \cite{lamontagne_discovery:2018,groves_scenarios:2007} or summarize a system's response to forcing \cite{Poff:2015jn,Steinschneider:2015kk,sriver_sealevel:2018}.
Despite the aversion to strong assumptions about the likelihood of different futures, subjective modeling decisions such as the choice of system model, the set of candidate decisions, the criteria used to assess outcomes, and the choice of how to sample \glspl{sow} strongly influence results \cite{quinn_exploratory:2020}.

@@

One way to interpret an ensemble of \glspl{sow} is as \gls{iid} draws from some true data generating process.
We call this scenario-conditional probabilistic analysis, which commonly arises when a probabilistically calibrated model for relevant outputs (\eg, a physical model for \gls{msl}; see \cref{sec:case-slr}) is forced by a deterministic scenario (\eg, a \gls{rcp} scenario).
In this case, \emph{the set of available \glspl{sow} is assumed to come from the true (often called population) distribution, and thus the sample statistics are assumed to reflect the true values.}
We add this concept to our framework with boxes (d) and (e) of \cref{fig:flowchart}, denoting the particular scenario (\ie, the assumed input) $M_k$.
If the \glspl{sow} are drawn \gls{iid} from $M_k$, the set of outcomes $u_{i, j}$ can be interpreted as \gls{iid} draws from the conditional distribution over outcomes, $p(u | x_i, M_k)$, allowing estimation of descriptive metrics.
This approach allows for quantification of uncertainty \emph{within models}, but can only qualitatively characterize uncertainty \emph{between models} \cite{wong_nola:2017,ruckert_coastal:2019,sharma_rcp:2021}. % Vivek raises an interesting point -- what if we incorporate the choice of model into the data generating process and use a posterior probability of each model being true -- but I think that's basically what we are calling step #3

@@

If the \gls{sow} ensemble is not drawn \gls{iid} from the true distribution (\eg, if one deliberately over-samples low-probability but high-impact \glspl{sow}) or if one wishes to synthesize \emph{across} multiple scenarios $M_1, M_2, \ldots, M_K$, then additional measures must be taken.
One possibility is to discard the \glspl{sow} ($\vb{s}$) and resample from some ``true'' distribution, which we shall denote $p_\mathrm{belief}$.
In practice, this is impractical because simulating $\vb{s}$ may rely on complicated models that are computationally expensive to re-run.

Instead of attempting to re-sample $\vb{s}$, we instead consider re-weighting.
A simple approach would be to assign each \gls{sow} $s_j$ a weight $w_j$ equivalent to the ratio of the desired (often called target distribution in survey analysis) distribution, $p_\mathrm{belief}$ and the sampling distribution, which could be approximated through methods such as kernel density estimation.
Paralleling joint probability methods for statistical analysis of tropical cyclones, we use a grid-based approach \cite{johnson_clara:2013,resio_probabilities:2007,toro_jpm-os:2010}.
First, we project the \glspl{sow} $\vb{s} \in \Omega$ onto a low-dimensional representation, which we denote $\qty{\psi_1, \psi_2, \ldots, \psi_J} \in \Psi$.
This allows us to specify a prior belief over this reduced space $p_\mathrm{belief}(\Psi)$ instead of over the full space $p_\mathrm{belief}(\Omega)$.
We then calculate a probabilistic weight $w_j \in \qty[0, 1]$ for each \gls{sow} $s_j$ so that the weighted distribution of $\psi_j$ closely approximates $p_\mathrm{belief}(\Psi)$.

We present here the case where the $\psi_j$ are one-dimensional.
We first sort the $\psi_j$  from least to greatest so that $\psi_{j-1} \leq \psi_j$, ($j \neq 1$).
Defining $F_\mathrm{belief}(s)$ to be the cumulative distribution corresponding to $p_\mathrm{belief}$, we calculate weights as
\begin{equation}\label{eq:weights}
  w_j = \begin{cases}
    F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_1 + \psi_2])                                                                     & j = 1     \\
    F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{j} + \psi_{j+1}]) - F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{j-1} + \psi_j]) & 1 < j < J \\
    1 - F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{J-1} + \psi_J])                                                             & j = J.
  \end{cases}
\end{equation}
This step is illustrated in \cref{fig:grid-sketch}.
For higher dimensional projections, this equation can be extended by partitioning the parameter space, then integrating the \gls{pdf} $p_\mathrm{belief}$ over each region.
From a practical perspective, many simulation models round outputs, meaning that there is a possibility of having $\psi_j$ that are equal; in this case it may be helpful to add a small amount of noise to the $\psi_j$ before computing the weights.
Diagnostic checks, such as examining the histogram of weights (not shown), may be valuable protections against degeneracy.

@@

The aim of this re-weighting framework is to integrate an ensemble of \glspl{sow} used for exploratory modeling into formal decision analysis, even when the \glspl{sow} deliberately over- or under-sample some regions of the parameter space.
As in \cref{sec:analysis-condition}, we must condition on a model: where the analysis of \cref{sec:analysis-condition} conditions upon deep uncertainties, the approach outlined in this subsection synthesizes across them.
We reiterate that stakeholders and experts will not, in general, agree on $p_\mathrm{belief}$ because its there is no ``true'' value of $p_\mathrm{belief}$ that could be objectively ascertained \cite{oreskes_verification:1994,walker_deep:2013}.
However, we posit that since we cannot be ``right,'' it is valuable to maximize the transparency of our implicit probabilistic assumptions, and suggest that writing down an explicit model for $p_\mathrm{belief}$ supports this aim.

@@

To illustrate the proposed decision analytic framework, we model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
We focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and approximate other model parameters as shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
  \item The decision vector $\vb{x}$ is comprised of discrete possible house heightenings ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$ f
  \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=70$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
  \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

@@

The choices of physical model and \gls{rcp} scenario jointly determine future \gls{msl} $p(\overline{y}|t)$.
\Cref{fig:lsl-evolution}(a) shows the time-varying 90\% credible intervals of \gls{msl} for three representative models.
The divergence between the the best-case (blue) and worst-case (red) models is small in the early 21st century and increases rapidly thereafter.
\Cref{fig:lsl-evolution}(b) shows the \glspl{pdf} of mean sea level in 2100 (dashed vertical line in panel (a)) under each of the sixteen models considered.
We return in \cref{sec:results-conditional} to the problem that these multiple models poses to decision makers.

@@

Other model validations lend confidence to the stationary \gls{gev} model selected.
For example, \cref{fig:surge-obs-return}(b) shows the estimated return periods for these storm surges; the estimated return period of the (shading) matches the (dots) the empirical plotting position dots.
Further, a positive control test (\cref{fig:surge-synthetic-data-experiment}) validates the model's ability to recover known parameter values.

@@

We now turn to the scenario-conditional analysis described in \cref{sec:analysis-condition}.
Whereas the exploratory analysis of the previous subsection interpreted each time series of future sea level as a sample from the space of possible futures, we can also interpret each \gls{sow} as a draw from one of the sixteen models of \gls{slr} shown in \cref{fig:lsl-evolution}.
As discussed in \cref{sec:analysis-condition}, this allows a formal estimation of decision metrics.

>>>>

\begin{keypoints}
  \item Decisions about how to manage climate risks are often made under deep uncertainty
  \item We introduce a Bayesian framework to transparently synthesize and characterize deep uncertainties with the goal to support decision-making
  \item We demonstrate the framework using a simple case study of house elevation for coastal flood risk management
\end{keypoints}

\begin{abstract}
  Projections of future climate risks can vary considerably from one source to another, posing considerable communication and decision-analytical challenges.
  One such challenge is how to present trade-offs under deep uncertainty in a salient and interpretable manner.
  Some common approaches include analyzing a small subset of projections or invoking Laplace's principle of insufficient reason to justify a simple average.
  These approaches can underestimate risks, hide deep uncertainties, and provide little insight into which assumptions drive decision-relevant outcomes.
  Here we introduce and demonstrate a transparent Bayesian framework for synthesizing deep uncertainties to inform climate risk management.
  The first step of this workflow is to generate an ensemble of simulations representing possible futures and analyze them through standard exploratory modeling techniques.
  Next, a small set of probability distributions representing subjective beliefs about the likelihood of possible futures is used to weight the scenarios.
  Finally, these weights are used to compute and characterize trade-offs, conduct robustness checks, and reveal implicit assumptions.
  We demonstrate the framework through a didactic case study analyzing how high to elevate a house to manage coastal flood risks.
\end{abstract}

@@

What are sound strategies to manage risks driven by climatic changes?
Addressing this question is complicated by the large uncertainties surrounding projections of the coupled natural-human systems that influence projections of climate risk.
These uncertainties often arise from choices experts have to make, for example about how to formulate scientific models of future water levels.
Different experts can disagree about these choices, leading to different plausible projections.
Analyzing decisions in such a situation of deep uncertainty poses nontrivial challenges.
At one extreme, picking a single representative projection can lead to under-estimation of risk and poor decisions.
At the other, communicating results separately for each projection can overwhelm decision-makers.
To make matters worse, typical approaches to this problem are mostly silent on what assumptions make a difference for the decisions at hand.
We develop and demonstrate a framework to address these challenges.
The framework provides a transparent approach to (i) combine a large number of deeply uncertain projections to a more understandable and smaller samples and (ii) provide insights as to how assumptions and modeling choices made by experts influence decisions.
We demonstrate the approach with a relatively simple example question of how high to elevate a house in the face of deeply uncertain projections of future water levels.

@@

Many critical infrastructure services around the world are aging and inadequate, and anticipated changes to regulation, economics and finance, patterns of population and infrastructure use, and climate challenge their ability to meet societal needs \cite{doss-gollin_txtreme:2021,doss-gollin_fatalism:2020,chester_reliable:2020,ho_dams:2017}.
The need to upgrade existing infrastructure and build new facilities motivates the question: for which possible futures should infrastructure systems and components be designed?
The answer to this question depends in part on values: intrinsic trade-offs between safety, performance, cost, and other objectives depend on context and stakeholder preferences \cite{keller_management:2021} and are often regulated by statute or industry guidelines.
For example, hospitals and critical infrastructure are generally designed to a higher standard of risk protection than ordinary buildings.
At the same time, performance depends upon future conditions, and so decisions about which scenario to design for are necessarily subject to implicit or explicit assumptions about the likelihood or possibility of different futures.

@@

Current practice in engineering, infrastructure design, and local governance relies heavily on standards that specify particular design events or conditions that buildings and infrastructure should safely withstand \cite{asce_7-10:2013,bruneau_multihazard:2017}.
These are often, though not always, informed by probabilistic analysis of relevant data.
For example, \gls{fema}, local governments, and engineering consultants produce local floodplain maps in many communities.
These trigger specific floodplain regulations, such as a requirement for homes in the 100-year floodplain with federally backed mortgages to be covered by flood insurance \cite{kousky_voucher:2014}.
Additionally, local building codes \cite<based on guidance such as>[]{asce_24-05:2006,FEMA_p-55:2011} may require new construction in flood zones to be elevated freeboard above a nominal \gls{bfe}.
Probabilistic analysis also informs the design of large-scale infrastructure.
For example, some levees in the Netherlands are designed for a nominal annual failure probability of $\nicefrac{1}{4000}$ \cite{eijgenraam_flooding:2014}, while a seawall proposed as part of a \$29 billion coastal protection project for  Galveston Bay was designed by setting the nominal annual probability of overtopping to 1\% \cite[Appendix D.,~p.~2-59]{USACE_coastal:2021}.

@@

The approahces currently used to estimate the \glspl{pdf} of climate hazards used in standard-setting and decision-making emphasize nominally objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies procedures for estimating flood frequency \cite{bulletin17c:2019} and \gls{noaa} Atlas 14 provides estimates of the intensity, duration, and frequency of extreme rainfall \cite{atlas14_texas:2018}.
Among several statistical assumptions these analyses make that have been recently called into question is stationarity \cite<the assumption that past and future hazard come from the same \gls{pdf}; see>[for a review]{merz_review:2014}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \cite{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}, and trends in many other hazards for other locations are consistent with observations \cite<see>[for a comprehensive summary]{ipcc_impacts:2022}.
While some methods have been proposed for incorporating trends into these analyses \cite<see>[for a review]{Salas:2018ge}, these assume specific forms of nonstationarity which may not fully represent physical processes or true levels of uncertainty \cite{DossGollin:2019,Montanari:2014hl,Serinaldi:2015bq}.
Ultimately, the difficulty of incorporating nonstationarity into existing frameworks has led to continued reliance on stationarity despite recognition of the associated limitations as discussed in p.~2 of \citeA{bulletin17c:2019} and p.~A.4-42 of \citeA{atlas14_texas:2018}.

@@

Although these methods for \gls{dmdu} have proven valuable in a wide range of settings, they still require necessarily subjective assessments about the likelihood of future conditions.
Even exploratory models require the analyst to choose which uncertainties are considered and how they are sampled.
When analyses use metrics that integrate performance over many possible futures, whether to compute sensitivities, estimate expectations, or compute robustness metrics, they necessarily make assumptions about the likelihood of different futures.
For example, uniform priors are often justified by deferring to Laplace's principle of insufficient reason \cite<see>[p.~135]{stigler_uncertainty:1986}, but even this choice makes strong assumptions that are sensitive to factors such as parameterization \cite[p.54]{gelman_bda3:2014}.
While one interpretation of deep uncertainty is that that specifying a joint \gls{pdf} over inputs is inappropriate, assumptions about the ranges and independence of parameters to sample are just as subjective as the choice of probability distribution \cite{schneider_scenarios:2002,quinn_exploratory:2020} and can often be interpreted  as a specific choice of probability distribution (we revisit this point in \cref{sec:analysis-condition}).
This motivates the development of decision analytic frameworks that draw from the strengths of \gls{dmdu} methods such as exploratory modeling, vulnerability assessment, robustness checks, and iterative stakeholder critique, but that embrace the reality that assumptions about the future are inescapable.

@@

We are motivated by parallels between decision making under deep uncertainty and the statistical problem of model selection.
\citeA{oreskes_verification:1994} argues that because natural systems are never closed and model results are never unique, validation and verification of models representing these systems is necessarily qualitative and subjective.
In the literature on decision making under deep uncertainty, this has led to recognition of the need to develop strategies that are robust to model imperfections.
We find parallels to this challenge in the Bayesian literature on model selection, particularly in the ``$\mathcal{M}$-open'' case in which there is no ``true'' model to identify and all models are imperfect representations of reality.
This is a theoretical contrast to the ``$\mathcal{M}$-closed'' case in which a true model can be identified and written down but is one amongst finitely many models from which an analyst has to choose \cite{bernardo_bayesian:1994}.
Commonly used classical methods for model selection implicitly assume $\mathcal{M}$-closed case and are defined by their ability to asymptotically identify the true model from a set of candidate models.
By contrast, model selection in the $\mathcal{M}$-open case emphasizes a subjective view of the modeling process, for which probability offers a self-consistent language with which to reason about the unknown rather than a statement of objective truth (see \citeA{gelman_philosophy:2013}, \citeA{ramsey_probability:2016}, or \citeA{jaynes_probability:2003} for a discussion of Bayesian philosophy and \citeA{Piironen:2017eh} for a technical discussion of methods for model selection).
From a practical perspective, model selection in the $\mathcal{M}$-open case emphasizes building models iteratively, simulating the consequences of these models, and subsequently using these simulations to critique and improve them until an acceptable model is reached \cite{gelman_workflow:2020}.

@@

We proceed as follows.
In \cref{sec:analysis} we present three formal decision analytic frameworks for analyzing an ensemble of \gls{slr} simulations, building through existing approaches for exploratory modeling (\cref{sec:analysis-explore}) and scenario analysis (\cref{sec:analysis-condition}) towards a formal Bayesian method for transparently synthesizing deep uncertainty through subjective prior beliefs (\cref{sec:analysis-synthesize}).
In \cref{sec:case-study} we describe the case study in detail.
Next in \cref{sec:results} we present results for each of the three decision lenses and discuss the advantages and limitations of each theoretical approach.
Finally in \cref{sec:conclusions} we discuss key findings, future research needs, and implications for policy and practice.

@@

A first analytical step is to use the model in an ``exploratory'' mode.
Exploratory modeling is averse to making explicit assumptions about the likelihood of different \glspl{sow} and instead seeks to generate new knowledge \cite{bankes:1993} by systematically exploring a large number of possible futures, emphasizing interactions between different uncertainties \cite{reed_msdbook:2022}.
Exploratory modeling is often paired with analyses that identify relevant scenarios  \cite{lamontagne_discovery:2018,groves_scenarios:2007} or summarize a system's response to forcing \cite{Poff:2015jn,Steinschneider:2015kk,sriver_sealevel:2018}.
Despite the aversion to strong assumptions about the likelihood of different futures, subjective modeling decisions such as the choice of system model, the set of candidate decisions, the criteria used to assess outcomes, and the choice of how to sample \glspl{sow} strongly influence results \cite{quinn_exploratory:2020}.

@@

One way to interpret an ensemble of \glspl{sow} is as \gls{iid} draws from some true data generating process.
This commonly arises when a single deep uncertainty (\eg, an emissions pathway) is used as an input for a fully probabilistic model; we call this scenario-conditional probabilistic analysis.
We add this concept to our framework with boxes (d) and (e) of \cref{fig:flowchart}, denoting the particular scenario (\ie, the assumed input) $M_k$.
If the \glspl{sow} are drawn \gls{iid} from $M_k$, the set of outcomes $u_{i, j}$ can be interpreted as \gls{iid} draws from the conditional distribution over outcomes, $p(u | x_i, M_k)$, allowing estimation of descriptive metrics.
This approach allows for quantification of uncertainty \emph{within} models of deeply uncertain processes, but can only qualitatively characterize uncertainty \emph{between} models \cite{wong_nola:2017,ruckert_coastal:2019,sharma_rcp:2021}.

@@

As discussed in the previous subsection, estimating the expectation of arbitrary functions under uncertainty is a critical component of decision making under uncertainty.
If the \glspl{sow} are drawn \gls{iid} from the true distribution, then the expected value of some function  $f$, $\mathbb{E}\qty[f(\vb{s})]$, can be readily estimated as $\frac{1}{N} \sum_{j=1}^N f(\vb{s}_j)$.
However, this is often not the case; for example, a fixed set of simulations may be available for analysis or low-probability but high-impact regions of the parameter space may have been sampled.
In this case, the formula must be adjusted to
\begin{equation}
  \mathbb{E}\qty[f(\vb{s})] \approx \frac{1}{N} \sum_{i=j}^N w_j f(\vb{s}_j),
\end{equation}
where the $w_j$ are suitably chosen weights such that their sum is equal to 1.

The challenge then becomes suitable estimation of the $w_j$.
Paralleling joint probability methods for statistical analysis of tropical cyclones, we use a grid-based approach \cite{johnson_clara:2013,resio_probabilities:2007,toro_jpm-os:2010}.
First, we project the \glspl{sow} $\vb{s} \in \Omega$ onto a low-dimensional representation, which we denote $\qty{\psi_1, \psi_2, \ldots, \psi_J} \in \Psi$.
Then, we partition the parameter space into a region corresponding to each \gls{sow} and integrating the \gls{pdf} $p_\mathrm{belief}$ over each region.

We present here the case where the $\psi_j$ are one-dimensional; extensions to higher dimensions are possible.
We first sort the $\psi_j$  from least to greatest so that $\psi_{j-1} \leq \psi_j$, ($j \neq 1$).
Defining $F_\mathrm{belief}(s)$ to be the cumulative distribution corresponding to $p_\mathrm{belief}$, we calculate weights as
\begin{equation}\label{eq:weights}
  w_j = \begin{cases}
    F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_1 + \psi_2])                                                                     & j = 1     \\
    F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{j} + \psi_{j+1}]) - F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{j-1} + \psi_j]) & 1 < j < J \\
    1 - F_\mathrm{belief}\qty(\frac{1}{2}\qty[\psi_{J-1} + \psi_J])                                                             & j = J.
  \end{cases}
\end{equation}
This step is illustrated in \cref{fig:grid-sketch}.
Diagnostic checks, such as examining the histogram of weights (not shown), may be valuable protections against degeneracy.

@@

The aim of this re-weighting framework is to integrate an ensemble of \glspl{sow} used for exploratory modeling into formal decision analysis, even when the \glspl{sow} deliberately over- or under-sample some regions of the parameter space.
As in \cref{sec:analysis-condition}, we must condition on a model: where the analysis of \cref{sec:analysis-condition} conditions upon deep uncertainties, the approach outlined in this subsection synthesizes across them.
We reiterate that stakeholders and experts will not, in general, agree on $p_\mathrm{belief}$ because there is not, even conceptually, an objectively correct choice \cite{oreskes_verification:1994,walker_deep:2013}.
However, we posit that since we cannot be ``right,'' it is valuable to maximize the transparency of our implicit probabilistic assumptions, and suggest that writing down an explicit model for $p_\mathrm{belief}$ supports this aim.

@@

To illustrate the proposed decision analytic framework, we model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
Following \citeA{zarekarizi_suboptimal:2020}, we focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and approximate other model parameters as shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
  \item The decision vector $\vb{x}$ is comprised of discrete possible house heightenings ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$.
  \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=70$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
  \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

@@

The choices of physical model and \gls{rcp} scenario jointly determine future \gls{msl} $p(\overline{y}|t)$.
\Cref{fig:lsl-evolution}(a) shows the time-varying 90\% credible intervals of \gls{msl} for three representative models.
The divergence between the the best-case (blue) and worst-case (red) models is small in the early 21st century and increases rapidly thereafter.
\Cref{fig:lsl-evolution}(b) shows the \glspl{pdf} of mean sea level in 2100 (dashed vertical line in panel (a)) under each of the sixteen models considered.
We return in \cref{sec:results-conditional} to the challenge of decision making under multiple \glspl{pdf}.

@@

Other model validations lend confidence to the stationary \gls{gev} model selected.
For example, \cref{fig:surge-obs-return}(b) shows the estimated return periods for these storm surges; the estimated return period  (shading) matches the the empirical plotting position (dots) and a positive control test (\cref{fig:surge-synthetic-data-experiment}) validates the model's ability to recover known parameter values.

@@

We now turn to the scenario-conditional analysis described in \cref{sec:analysis-condition}.
Whereas the exploratory analysis of the previous subsection interpreted each time series of future sea level as a sample from the space of possible futures, we can also interpret each \gls{sow} as a draw from one of the sixteen models of \gls{slr} shown in \cref{fig:lsl-evolution}.
As discussed in \cref{sec:analysis-condition}, this allows a formal estimation of decision metrics, conditional on the chosen scenario.
